\documentclass{article}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{color}
\usepackage{xspace}
\usepackage{url}

\title{On the input language of MinisatID}
\author{Broes De Cat}

\begin{document}

\maketitle

\section{Supported formats}
Several input formats are supported:
\begin{description}
	\item[opb] open pseudo-boolean competition format, defined in www.cril.univ-artois.fr/PB11/format.pdf
	\item[cnf] dimacs clausal normal form, defined in www.satcompetition.org/2011/rules.pdf
	\item[asp] ground answer set program format, defined in www.tcs.hut.fi/Software/smodels/lparse.ps.gz
	\item[qbf] quantified boolean formula format, defined in www.qbflib.org/Draft/qDimacs.ps.gz
	\item[ecnf] see the BNF grammar below
	\item[mecnf] experimental extension for modal operators (allows to express that, given a partial interpretation, a subtheory has a model / all extensions are models). No details will be provided yet in this document.
\end{description}

From an implementation point of view, the formats cnf, qbf, ecnf and mecnf are detected automatically. The formats opb and asp have to be specified by the option "{\verb --format=x }", with x="{\verb opb }", respectively x="{\verb asp }".

\section{ECNF format}
ECNF or Extended Clausal Normal Form is the native input language of the solver MinisatID.

\subsection{Syntax}
The syntax of the language can be expressed as the following BNF grammar ("+" expresses one or more, "*" expresses zero or more):

\begin{verbatim}
<formula>   ::= <comment>* "p ecnf" <header>* <EOL> <ecnftheory>*
<comment>   ::= "c" <char (not EOL)>* <EOL>
<header>    ::= "def" | "aggr" | "mnmz"

<ecnftheory>::= <comment>|
                <clause>|
                <rule1>|<rule2>|
                <set>|<wset>|<agg1>|<agg2>|
                <mnmz>|<subsetmnmz>|<aggmnmz>|
                <rangevar>|<enumvar>|<binrelvi>|<binrelvv>|
                <sumi>|<sumv>|sumvi>|<sumii>|<count>|<alldiff>

<clause>    ::= <lit>* <end>

<sem>       ::= "D"|"C"
<rule1>      ::= <sem> <atom> <lit>* <end>
<rule2>      ::= <sem> <defdelim> <id> <atom> <lit>* <end>

<set>       ::= "Set " <id> <lit>+ <end>
<wset>      ::= "WSet " <id> <tuple>+ <end>

<aggop>     ::= "L"|"G"
<aggtype>   ::= "Sum" | "Min" | "Max" | "Prod" | "Card"
<agg1>       ::= <aggtype> <sem> <aggop> <atom> <id> <int> <end>
<agg2>       ::= <aggtype> <defdelim> <id> <aggop> <atom> <id> <int> <end>

<mnmz>      ::= "Mnmlist " <lit>+ <end>
<subsetmnmz>::= "Mnmsubset " <lit>+ <end>
<aggmnmz>   ::= "Mnmagg " <aggtype> <atom> <id> <end>

<cpop>      ::= ">" | "<" | "=<" | ">=" | "=" | "~="
<rangevar>  ::="INTVARDOM" <var> <int>+ <end>
<enumvar>   ::="INTVAR" <var> <int> <int> <end>
<binrelvi>  ::="BINTRI"	<atom> <var> <cpop> <int> <end>
<binrelvv>  ::="BINTRT"	<atom> <var> <cpop> <var> <end>
<sumi>      ::="SUMSTRI" <atom> <var>+ <cpop> <int> <end>
<sumv>      ::="SUMSTRT" <atom> <var>+ <cpop> <var> <end>
<sumii>     ::="SUMSTSIRI" <atom> <vartuple>+ <cpop> <int> <end>
<sumvi>     ::="SUMSTSIRT" <atom> <vartuple>+ <cpop> <var> <end>
<count>     ::="COUNTSTRIT" <var>+ <op> <int> <var> <end>
<alldiff>   ::="DISTINCT" <var>+ <end>

<vartuple>    ::= "<"<var>","<int>">"
<tuple>     ::= <lit>"="<int>	

<atom>      ::= <posint>
<lit>       ::= "-"<posint>|<posint>
<id>        ::= <posint>
<var>       ::= <int>
<posint>    ::= any non-zero positive integer
<int>       ::= any integer

<defdelim>  ::= "<-"

<end>       ::= "0"
\end{verbatim}

Additional constraints on the language:
\begin{itemize}
	\item Each element is separated by at least one whitespace. Extra whitespaces or newlines can be added anywhere, only newlines within comments are not allowed.
	\item Any atom can only be the head of one rule or aggregate. \textbf{Exception}: An atom can occur as the head of multiple aggregate expressions, as long as there is only one such atom and there is a unit clause making it certainly true. This prevents inventing zillions of dummy atoms for e.g. large opb instances.
	\item Any id can only be defined once within the same type of statement.
	\item A set has to be defined before any aggregate using it.
	\item The header is provided for backwards-compatibility purposes, but is deprecated.
	\item 
\end{itemize}
	
\subsection{Semantics}

\paragraph{Literal} { \verb <lit> } a literal is an atom (positive integer) or its negation (negative integer)

\paragraph{Clause} 
\begin{verbatim}
<clause>    ::= <lit>* <end>
\end{verbatim} 
a clause is the disjunction of its literals 

\paragraph{Rule}
\begin{verbatim}
<sem>       ::= "D"|"C"
<rule>      ::= <sem> <atom> <lit>* <end>
<rule2>      ::= <sem> <defdelim> <id> <atom> <lit>* <end>
\end{verbatim} 
A rule defines its atom in terms of a disjunction (sem="D") or a conjunction (sem="C") of a set of literals. Any definition is evaluated according to the well-founded semantics (stable semantics can be specified as an option). In the case of rule1, all rules belong to the same definition. In the case of rule2, a definition id is specified, denoting the definition to which it belongs. If any rule is specified without a definition ID, no other rule can have a definition ID.

\paragraph{Sets}
\begin{verbatim}
<set>       ::= "Set " <id> <lit>+ <end>
<wset>      ::= "WSet " <id> <tuple>+ <end>
\end{verbatim} 
A set is identified by an id and is a non-empty set of tuples literal=integer. The first notation is an abbreviation of literal=1. It signifies that the literal has the associated weight if it is true, and does not belong to the set if it is false.

\paragraph{Aggregates}
\begin{verbatim}
<sem>       ::= "D"|"C"
<aggop>     ::= "L"|"G"
<aggtype>   ::= "Sum" | "Min" | "Max" | "Prod" | "Card"
<agg1>       ::= <aggtype> <sem> <aggop> <atom> <id> <int> <end>
<agg2>       ::= <aggtype> <defdelim> <id> <aggop> <atom> <id> <int> <end>
\end{verbatim} 
Given an example expression
\begin{verbatim}
WSet 1 1=2 -2=3 0
Sum D G 4 1 3 0
\end{verbatim}
The semantics is that the atom 4 is defined by the aggregate expression $sum(set_1) \geq 3$. 3 is called the \emph{bound}, 4 is the \emph{head}. If sem="C", completion semantics is used, if sem="D", well-founded semantics (with recursive aggregates) is used. 

Again, agg1 assumes all rules and defined aggregates are part of the same definition, agg2 defines a partition into multiple definitions. If agg1 is used, no agg2 or rule2 are allowed to occur in the theory
.
The available aggregate types are sum (also called pseudo-boolean / linear constraint), cardinality (count), maximum, minimum and product. The operator is either "G", $aggregate \geq bound$, or "L", $aggregate \leq bound$.
 
\paragraph{Optimization}
Optimization can be one of three types:
\begin{itemize}
\item Minimize ordered list ({ \verb <mnmz> }): given a list of literals, ordered from best (left) to worst, a model is better if it makes a better literal true.  
\item Subsetminimization ({ \verb <subsetmnmz> }): find the smallest subset of the set of literals for which a model of the theory exists.
\item Aggregate optimization ({ \verb <aggmnmz> }): minimize the value of the aggregate, given its type, its head and its set.
\end{itemize}

\paragraph{Constraints}
A number of finite domain constraints is support by the input format, probably subject to a serious extension in the near future.

\begin{verbatim}
<rangevar>  ::="INTVARDOM" <var> <int>+ <end>
<enumvar>   ::="INTVAR" <var> <int> <int> <end>
\end{verbatim} 

A finite-domain variable is identified by an identifier (var) and defined either by a range, specified by its minimum and maximum value (in that order and inclusive), or by an enumeration of all values it can map to.

\begin{verbatim}
<cpop>      ::= ">" | "<" | "=<" | ">=" | "=" | "~="
<binrelvi>  ::="BINTRI"	<atom> <var> <cpop> <int> <end>
<binrelvv>  ::="BINTRT"	<atom> <var> <cpop> <var> <end>
<sumi>      ::="SUMSTRI" <atom> <var>+ <cpop> <int> <end>
<sumv>      ::="SUMSTRT" <atom> <var>+ <cpop> <var> <end>
<sumii>     ::="SUMSTSIRI" <atom> <vartuple>+ <cpop> <int> <end>
<sumvi>     ::="SUMSTSIRT" <atom> <vartuple>+ <cpop> <var> <end>
<count>     ::="COUNTSTRIT" <var>+ <op> <int> <var> <end>
<alldiff>   ::="DISTINCT" <var>+ <end>
\end{verbatim} 

The currently supported constraints are (where $op$ is either $\leq, \geq, =, \neq, <, >$)
\begin{itemize}
\item binary relations of the form $var~op~var$ or $var~op~integer$, reified by a logical atom.
\item sum constraints of the form $sum(varlist)~op~int$, $sum(varlist)~op~var$, $sum(list(var\times int))~op~int$, $sum(list(var \times int))~op~var$, also reified by a logical atom.
\item a global count constraint of the form $count({var | varlist, var~op~int}) = var$.
\item a global alldifferent constraint of the form $alldiff(varlist)$.
\end{itemize}

\section{License}
Copyright 2007-2011 Katholieke Universiteit Leuven

Use of this software is governed by the GNU LGPLv3.0 license

Written by Broes De Cat and Maarten Mariën, K.U.Leuven, Departement Computerwetenschappen, Celestijnenlaan 200A, B-3001 Leuven, Belgium
\end{document}
